---
title: "BLS8 Scout-DSA Tx Summary Data"
author: "Salma Abdel-Raheem"
format: html
editor: visual
---

## BLS8

This is an overview of how I've compiled and processed the summary data I am presenting at the BLS8 conference in March 2024

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())
library(lubridate)
library(tidyverse)
```

## Transmitted Bio-logging Data

Processing transmitted (Tx) Data from Wildlife Computers Scout DSA tags that were deployed on 64 northern elephant seals (NES).

The following are metrics of interest for presentation:

Where do they go?

-   max latitude

-   max longitude

-   distributions of lat/long –\> if time

-   distance from colony (check if code works, omit if not)

-   distance from shore (probably not)

What do they do while they're there?

-   Dive Duration

    -   avg + max

-   dive depth

    -   avg + max dive depth

    -   frequency of dives per trip (box plot, 4 for each sex, group by age and color by sex) -\> nested box plot

    -   distribution of depth by age x sex (secondary)

-   Prey Capture Attempts

    -   per dive

    -   per day

-   Surface Interval

    -   Average across all dives

    -   Maximum across all dives

How does that affect fitness?

-   growth (mass gain, structural growth i.e. length)

-   survival

-   trip duration (arrive and depart/last obs dates)

#### Importing and tidying up Morphology + Tx data

```{r MoRphs data, message=FALSE, warning=FALSE}
morphs_full <- read_csv("data/Morphs_US_Crocker_RoDBLS8.csv", 
                        na = c("Pt.Reyes", "NA"))

#Trim to include only metadata, curve + straight lengths, and actual mass
morphs_trim <- morphs_full %>% 
  select(-8:-15, -18:-19, -matches("_girth"))

#Creating a df where each seal has one row for deployment AND recovery (if applicable)

#First create a df in which I filter out seals that were deployed AND recovered (RoD2022-n = 22, RoD2023-n = 21 * 2 for each season [deployment + recovery] = 86 rows)
morphs_repeats <- morphs_trim %>%
  group_by(topp_id) %>%
  filter(n() > 1)

# Pivot wider
morphs_pivot <- morphs_repeats %>%
  pivot_wider(names_from = tagging_procedure, values_from = 
                c(date, curve_length, standard_length, actual_mass))

#Let's create a new df that has information for animals that were NOT recovered, (RoD2022-n = 10, RoD2023-n = 11 = 21 rows) & format it to match the pivoted df 
morphs_depONLY <- morphs_trim %>%
  group_by(topp_id) %>%
  filter(n() == 1) %>% 
  #delete the procedure type column as that was "used up" when pivoting 
  select(-tagging_procedure) %>% 
  #move the date column to match the pivoted df
  relocate(date, .before = curve_length) %>% 
  #rename the date, lengths, and mass colums to reflect deployment measurements
  rename(date_Deployment = date, 
         curve_length_Deployment = curve_length,
         standard_length_Deployment = standard_length, 
         actual_mass_Deployment = actual_mass) %>% 
#create empty date, lengths, and mass columns and populate them with NAs to match     the structure of the pivoted df
  add_column(date_Recovery = NA, .after = "date_Deployment") %>% 
  add_column(curve_length_Recovery = NA, .after = "curve_length_Deployment") %>% 
  add_column(standard_length_Recovery = NA, .after = "standard_length_Deployment") %>%
  add_column(actual_mass_Recovery = NA, .after = "actual_mass_Deployment")

# Now combined the recovered and unrecovered seals dfs
morphs_combined <- bind_rows(morphs_pivot, morphs_depONLY) %>% 
  arrange(topp_id)

```

```{r Tx data, message=FALSE, warning=FALSE}
#Create a vector of all relevant data files and their paths
transmitted_dat <- dir('/Users/salmaabdel-raheem/Documents/GitHub/PhD/RoD/Transmitted/', 
                   pattern = '.*DSA.csv', 
                   recursive = TRUE, 
                   full.names = TRUE) 

#Merges all data files into one csv - no need for rbind!
transmitted_dat <- read_csv(transmitted_dat, id = "file") 

#Cleaning up the imported data & adding info like dive number, separating out
#DeployID into age, sex, and sealID columns
transmitted_f <- transmitted_dat %>% 
  #group by each individual seal
  group_by(DeployID) %>% 
  #create a column that starts counting at the start of each dive (segment == 0)
  mutate(test1 = case_when(Segment == 0 ~ 1), 
         #creates a column that counts and assigns a segment to a specific dive
         #based on the cumulative sum of the preceding value in the test1
         #counter & ignores NAs
         test2 = cumsum(replace_na(test1, 0)), 
         test3 = ifelse(is.na(Segment), NA, test2)) %>% 
  ungroup() %>% 
  #move those counter columns closer to the segment column to make it easier to
  #look at
  relocate(test1:test3, .after = Segment) %>% 
  #fill in the NA values with the dive number based on the value right below,
  #so it fills the data "up" instead of "down" which is the default
  fill(test3, .direction = "up") %>% 
  #renaming this counter column #commitment
  rename(DiveNum = test3) %>% 
  #deleting the counter columns and the DiveNumber column bc in this case it's
  #empty and is confusing to work with
  dplyr::select(-DiveNumber,-test1,-test2) %>%  
  #Separating animalID from age and sex and deleting the messy age_sex field
  separate(DeployID, c("animalid", "Age_Sex"), sep = " - ", remove = TRUE) %>% 
  select(-"Age_Sex") %>% 
  #Which seals were recovered -> survivorship metric
  mutate(Recovered = !grepl("-Offline", file),
         #Making Divestart into a Date-Time object
         DiveStart = parse_date_time(DiveStart, "%H:%M:%S %d-%b-%Y "),
         #Pulling data and time data into seperate columns
         Date = as.Date(DiveStart),
         Time = format(DiveStart, format = "%H:%M:%S"))%>% 
  #Tidying data frame to appear more intuitive
  relocate(Date:Time, .after = DiveStart) %>% 
  relocate(Recovered, .after = animalid) %>% 
  select(-file, -MobilityAvg, -MobilitySD, -Pitch:-TempDeep)
  
write_csv(transmitted_f, "data/transmitted_f.csv")
```

```{r}

```

```{r dive + morphology data}
all_dat <- left_join(morphs_combined, transmitted_f, by = "animalid")
```

#### Compiling dive segments to calculate per dive summary statistics

```{r}
#Calculating maximum dive depth, no minimum because that's just 0, and total dive
#duration for every dive in minutes
diveStat_perDive <- transmitted_f %>% 
  group_by(DeployID, DiveNum) %>% 
  summarise(
#The following values repeat for each dive segment in the raw data, so grabbing the first observation for each piece of data
    #Deployment Season
    DepSeason = first(DepSeason),
    #Seal ID
    SealID = first(SealID),
    #Was the tag recovered?
    Recovered = first(Recovered),
    #Age and Sex data (numerical and factored)
    Age = first(Age),
    Age_f = first(Age_f),
    Sex_f = first(Sex_f),
    #Date-time of dive start per dive per seal
    DateTime = first(DiveStart),
    Date = first(Date),
    Time = first(Time),
    #Location quality per dive per seal
    LocQual = first(LocationQuality),
    #Latitude per dive per seal
    Lat = first(Latitude),
    #Longitude per dive per seal
    Long = first(Longitude),
    #select the first row of a dive that has the total duration in it, and does
    #not have a segment assignment (hence, segment is.na) or you can do it as
    #Duration[1]
    #Dive duration divided by 60 to get minutes
    DiveDur_min = round(Duration[is.na(Segment)]/60, digits = 2), 
    #Maximum depth recorded during a dive
    MaxDiveDepth = max(InitialDepth, na.rm = TRUE), 
    #Average depth of a dive - ignore?
    MeanDiveDepth = mean(InitialDepth, na.rm = TRUE), 
    #Median depth of a dive - ignore?
    MedDiveDepth = median(InitialDepth, na.rm = TRUE),
    #ignore these behavior metrics?
    #total prey capture events per dive for each seal
    TotalPreyCap = sum(ActivityCount, na.rm = TRUE),
    #total swim effort per dive for each seal
    TotalSwim = sum(SwimEffort, na.rm = TRUE),
    #Maximum swim effort per dive for each seal
    MaxSwim = max(SwimEffort, na.rm = TRUE),
    #Average swim effort per dive for each seal
    MeanSwim = mean(SwimEffort, na.rm = TRUE)) %>% 
  ungroup()

write_csv(diveStat_perDive, "data/diveStat_perDive.csv")
```

##### Visualizing the distributions of some of the per-dive calculated metrics

```{r per dive histogRams , message=FALSE, warning=FALSE}
#Distribution of latitudes - look at east-west movement
Lat_hist <- ggplot(diveStat_perDive, aes(x = Lat, fill = Sex_f)) + 
  geom_histogram(width = 0.2, position = "dodge", binwidth = 10) +
  facet_wrap(.~Age_f, scales = "free_y") +
  ggtitle("Latitude by Age x Sex")
Lat_hist

#Distribution of longitude - look at north-south movement
Long_hist <- ggplot(diveStat_perDive, aes(x = Long, fill = Sex_f)) + 
  geom_histogram(width = 0.2, position = "dodge", binwidth = 10) +
  facet_wrap(.~Age_f, scales = "free_y") +
  ggtitle("Longitude by Age x Sex")
Long_hist

#Distribution of dive duration
dDur_hist<- ggplot(diveStat_perDive, aes(x = DiveDur_min, fill = Sex_f)) + 
  geom_histogram(width = 0.2, position = "dodge", binwidth = 10) +
  facet_wrap(.~Age_f, scales = "free_y") +
  ggtitle("Dive Duration by Age x Sex")
dDur_hist

#Distribution maximum dive depth
maxDDepth_hist<- ggplot(diveStat_perDive, aes(y = MaxDiveDepth, fill = Sex_f)) + 
  geom_histogram(width = 0.2, position = "dodge", binwidth = 10) +
  scale_y_reverse()+
  facet_wrap(.~Age_f, scales = "free_y") +
  ggtitle("Max Dive Depth by Age x Sex")
maxDDepth_hist

#Distribution of mean dive depth
meanDDepth_hist<- ggplot(diveStat_perDive, aes(y = MeanDiveDepth, fill = Sex_f)) + 
  geom_histogram(width = 0.2, position = "dodge", binwidth = 10) +
  scale_y_reverse()+
  facet_wrap(.~Age_f, scales = "free_y")+
  ggtitle("Mean Dive Depth by Age x Sex")
meanDDepth_hist

#Distribution of median dive depth
medDDepth_hist<- ggplot(diveStat_perDive, aes(y = MedDiveDepth, fill = Sex_f)) + 
  geom_histogram(width = 0.2, position = "dodge", binwidth = 10) +
  scale_y_reverse()+
  facet_wrap(.~Age_f, scales = "free_y") +
  ggtitle("Median Dive Depth by Age x Sex")
medDDepth_hist

#Distribution of max dive depth x duration by Age and Sex
dDur_maxDepth_plot <- ggplot(diveStat_perDive, 
                            aes(x = DiveDur_min, y = MaxDiveDepth, 
                                color = Sex_f)) + 
  geom_jitter() +
  scale_y_reverse()+
  facet_wrap(.~Age_f) +
  ggtitle("Dive Duration by Depth") +
  geom_smooth(method = "lm")

dDur_maxDepth_plot
```

Getting some weird `NA` values in `Age_f` and `Sex_f` fields. Let's try to figure it out...

```{r debugging a non-error bug...per dive statistics}
# Check for NA values in Age_f variable
na_age <- diveStat_perDive %>%
  filter(is.na(Age_f))

# Check for NA values in the Sex_f variable
na_sex <- diveStat_perDive %>%
  filter(is.na(Sex_f))

# Count NA values in each variable
count_na <- diveStat_perDive %>%
  summarise(across(c(Age_f, Sex_f), ~ sum(is.na(.))))

# Print the results
print(na_age)
print(na_sex)
print(count_na)
```

Hmmm...so the per dive statistics dropped some of the age and sex values for some of the seals, not sure where this happened so lets see if these `NA` values manifested in the original `transmitted_f` compiled data frame.

```{r debugging a non-error bug...compiled transmitted data}
#Check for NA values in Age_f variable
na_age <- transmitted_f %>%
  filter(is.na(Age_f))

# Check for NA values in the Sex_f variable
na_sex <- transmitted_f %>%
  filter(is.na(Sex_f))

# Count NA values in each variable
count_na <- transmitted_f %>%
  summarise(across(c(Age_f, Sex_f), ~ sum(is.na(.))))

# Print the results
print(na_age)
print(na_sex)
print(count_na)
```

Okay so this seemed to happen when `Age` and `Sex` were made into factored variables initially. Let's see if the originally formatted `Age` and `Sex` data were split up correctly when compiling `transmitted_f`.

```{r debugging a non-error bug... non-factored age and sex data}
# Check for NA values in Age_f variable
na_age <- transmitted_f %>%
  filter(is.na(Age))

# Check for NA values in the Sex_f variable
na_sex <- transmitted_f %>%
  filter(is.na(Sex))

# Count NA values in each variable
count_na <- transmitted_f %>%
  summarise(across(c(Age, Sex), ~ sum(is.na(.))))

# Print the results
print(na_age)
print(na_sex)
print(count_na)
```

Something weird happened when parsing age and sex information from the original `DeployID` column. Let's look at `SealID == K335`

```{r}
```
